---
title: "Analyzing Water Insecurity with DHS Data in R"
author: "Jesse McDevitt-Irwin"
format: html
editor: visual
---

## 🧭 Overview

In this lab, we will:

-   Clean and recode variables for the JMP Drinking Water Ladder
-   Construct the 12-item WISE (Water InSecurity Experiences) Scale
-   Visualize patterns of water access and insecurity using `ggplot2`
-   Practice using `mutate()`, `case_when()`, `across()`, and `summarize()`

We'll use the **2022-23 Mozambique DHS** household dataset. You should have it loaded as `hr_raw`.

------------------------------------------------------------------------

## 🧪 Setup

```{r}
# Load required packages
library(tidyverse)
library(haven)
library(summarytools)

# Clear your R environment
rm(list = ls())

# Select only relevant household variables
vars_hr <- c(
  "hhid", "hv001", "hv002", "hv005", "hv008", "hv009", "hv024", "hv025", "hv204",
  "shwa17","hv025",
  "sh108g", "sh108o", "sh108h", "sh108i", "sh108j", "sh108k",
  "sh108l", "sh108m", "sh108n", "sh108p", "sh108q", "sh108r"
)

# Load the household recode file
hr_raw <- read_dta("Raw/MZHR81FL.DTA", col_select = all_of(vars_hr))
```

------------------------------------------------------------------------

## 🔹 JMP Drinking Water Ladder

### Step 1: Classify JMP levels

```{r}
# Fill in variable names and categories
hr_raw <- hr_raw %>%
    mutate(jmp_water = case_when(
    is.na(shwa17) ~ NA,
    shwa17 %in% c(10,11) ~ "Safely Managed",  # piped into dwelling
    shwa17 %in% c(12,13,14,21,31,41) & hv204 <= 30 ~ "Basic",
    shwa17 %in% c(12,13,14,21,31,41) & hv204 > 30 ~ "Limited",
    shwa17 %in% c(32,42,51,71,72) ~ "Unimproved",
    shwa17 == 81 ~ "Surface Water",
    TRUE ~ "Other/Unknown"
  )) %>% 
   mutate(jmp_water = factor(jmp_water,
                            levels = c("Safely Managed", 
                                       "Basic", "Limited", 
                                       "Unimproved", "Surface Water", 
                                       "Other/Unknown"),
                            ordered = TRUE))

```

**Think-Pair-Share**: What is the overall purpose of this code? Look through each line. Find at least one line you do not understand. Discuss with your neighbor.

### Step 2: Plot JMP ladder

```{r}
# Plot a bar chart of JMP categories
hr_raw %>% drop_na(jmp_water) %>% 
  ggplot(aes(x = jmp_water)) +
  geom_bar() +
  coord_flip() +
  labs(title = "JMP Drinking Water Ladder", x = "", y = "Households")
```

**Think-Pair-Share**: What is something you'd like to change about this figure, visually?

------------------------------------------------------------------------

## 🔹 WISE Scale (12 Items)

### Step 1: Recode 1–5 values to 0–3

```{r}
wise_vars <- c("sh108g", "sh108o", "sh108h", "sh108i", "sh108j", "sh108k",
               "sh108l", "sh108m", "sh108n", "sh108p", "sh108q", "sh108r")

# Use case_when to recode
hr_raw <- hr_raw %>%
  mutate(across(all_of(wise_vars), ~ case_when(
    .x == 1 ~ 0,
    .x == 2 ~ 1,
    .x == 3 ~ 2,
    .x %in% c(4,5) ~ 3,
    TRUE ~ NA_real_
  ), .names = "recoded_{.col}"))
```

### Step 2: Calculate WISE score

```{r}
hr_raw <- hr_raw %>%
  mutate(WISE_score = rowSums(across(starts_with("recoded_")))) %>% 
  mutate(WISE_category = case_when(
            WISE_score <= 2 ~ "No-to-marginal",
            WISE_score <= 11 ~ "Low",
            WISE_score <= 23 ~ "Moderate",
            WISE_score >= 24 ~ "High",
            TRUE ~ NA)
        ) %>% 
  mutate(WISE_category = factor(WISE_category,
                                levels = c("No-to-marginal", "Low", 
                                           "Moderate", "High"),
                                ordered = TRUE))
```

**Think-Pair-Share**: What is the overall purpose of these code chunks? Look through each line. Find at least one line you do not understand. Discuss with your neighbor.

### Step 3: Plot WISE score distribution

```{r}
hr_raw %>%  
  ggplot(aes(x = WISE_category)) +
  geom_bar() +
  coord_flip() +
  labs(title = "WISE Score Distribution", x = "Score (0–36)", y = "Households")
```

**Think-Pair-Share**: What is something you'd like to change about this figure, visually?

------------------------------------------------------------------------

### Step 4: WISE Score by Urban-Rural

```{r}
hr_raw %>%
  mutate(hv025 = as_factor(hv025)) %>% 
  group_by(hv025) %>%
  summarise(mean_wise = mean(WISE_score)) %>%
  ggplot(aes(x = hv025, y = mean_wise, fill = hv025)) +
  geom_col() +
  labs(title = "Average WISE Score by Residence Type",
       x = "Residence Type",
       y = "Average WISE Score") +
  theme_minimal()
```

**Think-Pair-Share**: What is something you'd like to change about this figure, visually?

## ✅ Bonus Challenge

-   Group WISE scores by wealth quintile (`hv270)` and plot with `ggplot`
